Maximum Entropy Spectral Analysis: a case study

### 2021-03-25 paper/mesa.tex
Modified theoretical foundations

\subsection{Maximum Entropy Principle} \label{sec:MAXENT}

\sschmidt{This section might be simplified...}

Before introducing MAXENT as a variational problem, we will workout some simple examples to contexualise how information's entropy
and `uncertainty' are related. First of all,  it is worth noting that the word `information' will not be used to indicate our knowledge on the system under study: this is what we will call `evidence'. Information will be intended as in communication theory: it specifies the length of the message necessary to provide
a full description of the system under study.

For instance, consider a perfectly sinusoidal signal: knowledge of amplitude, frequency and phase are sufficient to fully reproduce it: 
it has a low information entropy content. Even less information is required if we are studying a system whose outcome is 
certain (has probability $p = 1$), as in this case, a communication is not even needed.  
Shannon \cite{Shannon} proposed the quantity
\begin{equation}\label{eq:information}
    I = \log_2 \frac{1}{p(x)}
\end{equation}
to represent the quantity of information brought by an outcome $x$ with probability $p(x)$. It is additive quantity as well as monotonically decreasing as a function of $p \in [0, 1]$. The more uncertain the outcome, the higher the information it brings.

We now seek to generalize the measure for the information in the case where different outcomes are possible.
To do so, let us consider an experiment with two different random outcomes $E_1, E_2$, with given probabilities $P_1$ and $P_2$, and we ask ourselfs which are the probability assignments that makes the outcome more uncertain (i.e. maximize the information).
If $P_1$ and $P_2$ are largely different, for istance $P_1 = 0.999$ and $P_2 = 0.001$, we clearly have enough evidence to believe that event $E_1$ will occur almost certainly, considering $E_2$ to be a very implausible outcome. The information content will be very low.
On the other hand, most unpredictable experiment happens when 
\begin{equation}\nonumber
    P_1 = P_2 = \frac{1}{2}:
\end{equation}
this describes a situation of `maximum ignorance'; we have no evidence at all and we have no way of predicting what the outcome of such an experiment will be.  The information content of this system must be high.

Of course, when assigning probabilities for a set of events, all the evidence available must be taken into account. On the other hand, the probability assignment must not introduce additional evidence in the system and thus the need to be ``the most uncertain possible given the knoledge of the system".

%In this context, the role played by evidence should be clear: it dictates the probability assignment in order to make the result `more certain' or `less informative'. Any assignment different from the `most uncertain' is violating the principle of it to being non committal to non-available evidence. We can only assign a more certain probability, i.e. less informative, if we have evidence for it. If we have none, the result has to be the most unpredictable, i.e. the most informative. From the point of view of communication theory: the higher the evidence on a system, the lower the information I need to send to describe it.
\par
MAXENT puts the aforementioned principle into a more precise mathematical context by stating that probabilities should be assigned by maximizing uncertainty using evidence as constraint. 
This defines a variational problem, where a numerical measure of uncertainty $H\left[p_1, \dots, p_N\right]$ has to maximized. $H$ has to be a `generalization' of information defined in equation \eqref{eq:information} to be something like the `expected information' brought by an experiment with $N$ possible outcomes each with its own probability $p_i$.
%
%Some properties for $H$ can be worked out easily. 
%From previous discussion, given a fixed number $N$ of possible outcomes, we want $H$ to be maximized by the choice $p_1 = \dots = p_N$.
%Consider now the opposite situation, in which probability is assigned appealing to IP and we have a different number of possible outcomes.
%Intuitively, the most unpredictable experiment is the one with larger number of outcomes: probability of "getting it wrong" increases as $\frac{N -1}{N}$. So when equal probabilities are assigned, we want
%\begin{equation}
%    \nonumber
%    H\left[\frac{1}{N}, \dots, \frac{1}{N}\right]
%\end{equation}
%to increase monotonically with N.
%
%We also want our function to be continuous in its parameters: any small change in probability has to slightly affect uncertainty. \\
%Last is composition property. Consider three events with probability $p_1, p_2, p_3$ such that $p_2 + p_3 = 1 - p_1 = q$. We require 
%\begin{equation}
%    H\left(p_1, p_2, p_3\right) = H(p_1, q) + qH\left(\frac{p_2}{q}, \frac{p_3}{q}\right),
%\end{equation}
%so that uncertainty is additive.

@@

>>>>

\subsection{Maximum Entropy Principle} \label{sec:MAXENT}
Before introducing MAXENT principle and its formulation, we will workout some simple examples to specify what the core quantities of the problem are and the 
role they play. These are the 'evidence' and the 'information'. Let us start with the 'information' for one or more events: it is a measure of the degree of uncertainty on the outcomes of some experiment and specifies the length of the message necessary to provide a full description of the system under study. For instance, consider a perfectly sinusoidal signal: knowledge of amplitude, frequency and phase are sufficient to fully reproduce it: 
it has a low information entropy content. Even less information is required if we are studying a system whose outcome is 
certain (has probability $p = 1$), as in this case, a communication is not even needed.  
Shannon \cite{Shannon} proposed the quantity
\begin{equation}\label{eq:information}
    I = \log_2 \frac{1}{p(x)}
\end{equation}
to represent the quantity of information brought by an outcome $x$ with probability $p(x)$. It is additive quantity as well as monotonically decreasing as a function of $p \in [0, 1]$: the more uncertain the outcome, the higher the information it brings.
We now seek to generalize the measure for the information in the case where different outcomes are possible.
To do so, let us consider an experiment with two different random outcomes $E_1, E_2$, with given probabilities $P_1$ and $P_2$, and we ask ourselfs which are the probability assignments that makes the outcome more uncertain (i.e. maximize the information).
If $P_1$ and $P_2$ are largely different, for istance $P_1 = 0.999$ and $P_2 = 0.001$, we clearly have enough evidence to believe that event $E_1$ will occur almost certainly, considering $E_2$ to be a very implausible outcome. The information content will be very low.
On the other hand, most unpredictable experiment happens when 
\begin{equation}\nonumber
    P_1 = P_2 = \frac{1}{2}:
\end{equation}
this describes a situation of `maximum ignorance':  the information content of this system must be high. If an experiments with $N$ outcomes
is  considered, the most uncertain experiment is the one in which 
\begin{equation}\label{eq:indifference}
P_1 = \dots = P_N = \frac{1}{N}.
\end{equation}
In the above cases our knowledge on the system was the total number of different outcomes. Let us now reverse the perspective: what is the most unbiased probability assignment we can make if all we know is that one over $N$ outcomes will occur? Since we have no reason to prefere one outcome over the others, the least biased assignment is the one equivalent to (\ref{eq:indifference}) and is found to be coincident with the most uncertain experiment. This result represents the core idea of MAXENT principle: the least biased probability assignment can be obtained maximizing the "expected" information's entropy. \\ 
In the previous example we had no reason to prefere one outcome over the others but that's not always the case. This hypothesis can be relaxed introducing the concept of 'evidence', that represents the knowledge we posses on the system under study. With proper evidence some outcomes can be preferred to others. Probability assignment must obey to all the evidence we possess. 
\par
MAXENT puts the aforementioned principle into a more precise mathematical context by stating that probabilities should be assigned by maximizing uncertainty using evidence as constraint. 
This defines a variational problem, where a numerical measure of uncertainty $H\left[p_1, \dots, p_N\right]$ has to maximized. $H$ has to be a `generalization' of information defined in equation \ref{eq:information} to be something like the `expected information' brought by an experiment with $N$ possible outcomes each with its own probability $p_i$.

@@

The maximisation of entropy, supplemented by evidence in the form of constraints to which the sought-for probability distribution must obey,
gives rise to several of the most common probability distributions utilised commonly by data scientists. For instance, whenever the only constraint available is the normalisation of the probability distribution, the entropy is maximised by the uniform distribution. 
If the information available is instead the expected value -- the mean -- entropy is maximised by the  exponential distribution
